---
title: "Section 5.2 - Cross-Validation, Ridge, Lasso"
from: markdown+emoji
format:
  html:
    toc: true
    toc-location: left
page-layout: full
description: |
author: Jacob Jameson

---

## About Dataset
### Context

This dataset is originally from the National Institute of Diabetes and Digestive and Kidney Diseases. The objective of the dataset is to diagnostically predict whether or not a patient has diabetes, based on certain diagnostic measurements included in the dataset. Several constraints were placed on the selection of these instances from a larger database. In particular, all patients here are females at least 21 years old of Pima Indian heritage.

### Content

The datasets consists of several medical predictor variables and one target variable, Outcome. Predictor variables includes the number of pregnancies the patient has had, their BMI, insulin level, age, and so on.

[Download the dataset here](https://github.com/jacobjameson/jacobjameson.github.io/blob/main/api%20222%20files/section%205/diabetes.csv):


```{r}
diabetes_data <- read.csv("diabetes.csv", header = TRUE)
```

Let's get to know our data. 

```{r}
dim(diabetes_data)
summary(diabetes_data)
```

From the output, we see that we have one binary variable (Outcome) and eight continuous variables. We have 768 observations.

Today, we will focus on predicting Outcome: (whether or not a patient has diabetes)

Let's start with KNN for classification, so load the required package:

```{r}
library(class)
```

We are interested in using KNN on this dataset, but we aren't sure which k will give the best out-of-sample accuracy. We can use cross-validation to find the best k in this sense.

### Brief intro to functions in R
An R function is created by using the keyword function. The basic syntax of an R function definition is as follows −

```{r, eval=F}
function_name <- function(arg_1, arg_2, ...) {
   Function body 
}
```

The different parts of a function are:

- Function Name: This is the actual name of the function. It is stored in R environment as an object with this name.

- Arguments: An argument is a placeholder. When a function is invoked, you pass a value to the argument. Arguments are optional; that is, a function may contain no arguments. Also arguments can have default values.

- Function Body: The function body contains a collection of statements that defines what the function does.

- Return Value: The return value of a function is the last expression in the function body to be evaluated.


```{r}
## Example of a simple function
add_fun <- function(a1, a2){
  res = a1 + a2
  return(res)
}

add_fun(3, 4)
```

### Function to perform cross-validation

Let's write a function to do cross-validation. We will name the function cross_validation_KNN() and it will take 3 arguments:

1. The feature columns of the data
2. The outcome column of the data (Y)
3. A vector of all k values to test
4. The number of folds to use in k-fold cross validation


```{r}
cross_validation_KNN  <- function(data_x, data_y, k_seq, kfolds) {
    
  ## We will start by assigning each observation to one and 
  ## only one fold for cross-validation. To do this, we use
  ## an index vector. The vector's length equals the number
  ## of observations in the data. The vector's entries are 
  ## equal numbers of 1s, 2s, etc. up to the number of folds
  ## being used for k-fold cross validation.
  ## Recall seq(5) is a vector (1, 2, 3, 4, 5)
  ## ceiling() rounds a number up to the nearest integer
  ## rep(a, b) repeats a b times (e.g. rep(1, 3) => (1, 1, 1))
  ## So this says repeat the sequence from 1:kfolds
  ## enough times to fill a vector that is as long or 
  ## slightly longer than the number of observations in
  ## the data. Then, if the length of the vector exceeds
  ## the number of observations, truncate it to the right
  ## length
  
  fold_ids <- rep(seq(kfolds), ceiling(nrow(data_x) / kfolds))
  fold_ids <- fold_ids[1:nrow(data_x)]
  
  ## To make the IDs random, randomly rearrange the vector
  fold_ids <- sample(fold_ids, length(fold_ids))
  
  ## In order to store the prediction performance for each fold
  ## for each k, we initialize a matrix.
  CV_error_mtx  <- matrix(0, nrow = length(k_seq), ncol = kfolds)
  
  ## To run CV, we will loop over all values of k that we want to 
  ## consider for KNN. For each value of k, we will loop over all
  ## folds. For each fold, we will estimate KNN for the given k on 
  ## all but one fold. We will then measure the model's accuracy 
  ## on the hold out fold and save it. After we finish looping 
  ## through all k's and all folds, we will find the average CV
  ## error for each value of k and use that as a measure of the 
  ## model's performance.
  for (k in k_seq) {
    for (fold in 1:kfolds) {
      ## Train the KNN model (Note: if it throws a weird error, make sure
      ## all features are numeric variables -- not factors)
      ## Note: usually the features are normalized/re-scaled
      ## Otherwise KNN will give more weight to variables at the larger scale
      ## when calculating the distance metric.
      ## See page 165 of the textbook for a useful example
      knn_fold_model <- knn(train = scale(data_x[which(fold_ids != fold),]),
                            test = scale(data_x[which(fold_ids == fold),]),
                            cl = data_y[which(fold_ids != fold)],
                            k = k)
      
      ## Measure and save error rate (% wrong)
      CV_error_mtx[k, fold] <- mean(knn_fold_model != 
                                      data_y[which(fold_ids == fold)])
    }
  }
  
  ## We want our function to return the accuracy matrix
  return(CV_error_mtx)
    
}
```

### Applying the function

We can now use the cross-validation function on real data. Note that the outcome is stored in the 9th column.

```{r}
set.seed(222)
knn_cv_error5 <- cross_validation_KNN(data_x = diabetes_data[, -9],
                                      data_y = diabetes_data$Outcome,
                                      k_seq = seq(20),
                                      kfolds = 5)
print(knn_cv_error5)

knn_cv_error10 <- cross_validation_KNN(data_x = diabetes_data[,-9],
                                       data_y = diabetes_data$Outcome,
                                       k_seq = seq(50),
                                       kfolds = 10)
print(knn_cv_error10)
```

We are usually interested in the mean CV error or accuracy for each parameter considered (value of k in this case). To find the mean for each row, we can use the function `rowMeans()`. Note that an equivalent function exists for taking the means of columns: `colMeans()`. There are also functions for `rowSums()` and `colSums()`.

```{r}
mean_cv_error5 <- rowMeans(knn_cv_error5)
mean_cv_error10 <- rowMeans(knn_cv_error10)
```

To find the position of the smallest CV error, use `which.min()`.

```{r}
which.min(mean_cv_error5)
which.min(mean_cv_error10)
```

To understand how perform changes across the K values, a plot would be helpful.

```{r}
plot(seq(20), mean_cv_error5, type = "l",
     main = "Mean CV Error Rate as a Function of k",
     ylab = "CV Error Rate", xlab = "k")
```


From this, we see that low k values have poor performance while large k values seem to have roughly equivalent performance.

## BONUS MATERIAL

### LDA/QDA/Logistic CV

Given that our problem is a classification problem, we might be interested in seeing how well we could do with different methods, like logistic regression, LDA and QDA. Let's see! We can edit the function we used for CV for KNN and make it relevant for these models. Note that `logistic_formula = NULL` means the function can run even if this formula is not specified (which it won't be when we want to run LDA or QDA). Note also that we need to load the package `MASS`, because it contains the functions `lda()` and `qda()`.

```{r}
library(MASS)
cross_validation <- function(full_data, model_type, kfolds,
                             logistic_formula = NULL) {
  
  ## Define fold_ids in exactly the same way as before
  fold_ids <- rep(seq(kfolds), 
                       ceiling(nrow(full_data) / kfolds))
  fold_ids <- fold_ids[1:nrow(full_data)]
  fold_ids <- sample(fold_ids, length(fold_ids))
  
  ## Initialize a vector to store CV error
  CV_error_vec  <- vector(length = kfolds, mode = "numeric")
  
  ## Loop through the folds
  for (k in 1:kfolds){
    if (model_type == "logistic") {
      logistic_model <- glm(logistic_formula,
                            data = full_data[which(fold_ids != k),],
                            family = binomial)
      logistic_pred <- predict(logistic_model,
                               full_data[which(fold_ids == k),],
                               type = "response")
      class_pred <- as.numeric(logistic_pred > 0.5)
      
    } else if (model_type == "LDA") {
      lda_model <- lda(full_data[which(fold_ids != k),-9], 
                       full_data[which(fold_ids != k),9])
      lda_pred <- predict(lda_model, full_data[which(fold_ids == k), -9])
      class_pred <- lda_pred$class
      
    } else if (model_type == "QDA") {
      qda_model <- qda(full_data[which(fold_ids != k), -9], 
                       full_data[which(fold_ids != k), 9])
      qda_pred <- predict(qda_model, 
                          full_data[which(fold_ids == k), -9])
      class_pred <- qda_pred$class
    }
    
    CV_error_vec[k] <- mean(class_pred != full_data[which(fold_ids == k), 9])
  }
  return(CV_error_vec)
}

## Run CV for logistic regression:
logistic_formula <- paste("Outcome", paste(colnames(diabetes_data)[1:8],
                                           collapse = " + "), 
                          sep = " ~ ")
logistic_CV_error <- cross_validation(diabetes_data, "logistic", 5, 
                                      as.formula(logistic_formula))

## Run CV for LDA:
lda_CV_error <- cross_validation(diabetes_data, "LDA", 5)

## Run CV for QDA:
qda_CV_error <- cross_validation(diabetes_data, "QDA", 5)

## Determine the best model in terms of lowest average CV error
print(paste("Logistic Error Rate:", round(mean(logistic_CV_error), 3)))
print(paste("LDA Error Rate:", round(mean(lda_CV_error), 3)))
print(paste("QDA Error Rate:", round(mean(qda_CV_error), 3)))
```
