---
title: "Section 2 - KNN and Linear Regression"
from: markdown+emoji
format:
  html:
    toc: true
    toc-location: left
page-layout: full
description: |
author: Jacob Jameson

---

# Notes 

Note that the material in these notes draws on past TF’s notes (Ibou Dieye, Laura Morris, Emily Mower, Amy Wickett), and the more thorough treatment of these topics in *Introduction to Statistical Learning* by Gareth James, Daniela Witten, Trevor Hastie and Robert Tibshirani.


## Important Machine Learning Concepts

### Regression vs Classification
Prediction problems can be defined based on the characteristics of the outcome variable we want to predict. 

- **Regression** problems are those where the outcome is quantitative 

- **Classification** problems are those where the outcome is qualitative / categorical

Sometimes the same methods can be used for regression and classification problems, but many methods are useful for only one of the two problem types.

### Bias-Variance Trade-off
The **variance** of a statistical learning method is the amount by which the prediction function would change if it was estimated on a different training set. A model that overfits has high variance, whereas a model that underfits has low variance. 

To remember the difference between low variance and high variance models, I find it helpful to think of examples. Suppose your model was ``use the mean of the training data as the predicted value for all new data points.'' The mean shouldn't change much across training sets, so this has low variance. On the other hand, a model that picked up super complex patterns is likely to be picking up noise in addition to signal. The noise will vary by training set, so such a method would have high variance. 

The **bias** of a statistical learning method is the error produced by representing a real world problem by a statistical learning method. Very flexible models (which are prone to overfitting) can capture complex patterns and so tend to have low bias. Very simple models (which are prone to underfitting) are limited in their ability to pick up patterns and so may have high bias. 

The book uses the example of representing a non-linear function by a linear one to show that no matter how much data you have, a linear model will not do a great prediction job when the process generating the data is non-linear. Bias also applies to methods that might not fit your traditional concept of a statistical function. In the K-Nearest Neighbors section, we will discuss bias in that setting. 

Often, we will talk about the **bias-variance trade-off**. In an ideal world, we would find a model that has low variance and low bias, because that would yield a good and consistent model. In practice, you usually have to allow bias to increase in order to decrease variance and vice versa. However, there are many models that will decrease one (bias or variance) significantly while only increasing the other a little. 

### Supervised v. Unsupervised Learning

**Supervised learning** refers to problems where there is a known outcome. In these problems, you can train a model to take features and predict the known outcome. 

**Unsupervised learning** refers to problems where you are interested in uncovering patterns and do not have a target outcome in mind. 

An example of supervised learning would be using students' high school grades, class enrollments, and demographic variables to predict whether or not they attend college. 

An example of unsupervised learning would be using the same grades, enrollment, and demographic features to identify ``types'' of high school students. That is, students who look similar according to these features. Perhaps you are interested in this because you want to make classes that contain a mix of different types of students. Often, unsupervised learning is useful for creating features for supervised learning problems, but sometimes uncovering patterns is the final objective.

### Measuring Model Performance

There are different functions you can use to measure model performance, and which function you choose depends on your data and your objective. These functions are called ``loss functions,'' which is a somewhat intuitive name when you think about the fact that your machine learning algorithm is trying to minimize this function and thus minimize your loss. 

To understand how and why loss functions depend on your data and objectives, examples can be helpful. 

Consider first that you are trying to predict the future college majors of this year's incoming freshmen (a classification problem). In this case, your prediction will either be right (you predict the major they end up choosing) or it will be wrong. Therefore, you might use accuracy (\% correct) to measure model performance. 

What if, though, you cared more about being wrong for some majors than others? For example, imagine that all biology majors are going to need personalized lab equipment in their junior year and that the lab equipment is really expensive if ordered last minute but a lot cheaper if ordered a year or more in advance? Then, you might want to give more weight to people who end up being biology majors so that your model does better for predicting biology majors than other majors.

Now consider that you are trying to predict home prices (a regression problem). You might measure your performance using mean-squared error (MSE), which is found by taking the difference between the predicted sale price for each home and the true sale price (the error), squaring it for each home, and then taking the mean of these squared errors. However, home prices are skewed (e.g. some homes are extremely expensive compared to most homes on the market). This means that a 5\% error on a \$3 million home is a lot bigger than a 5\% error on a \$100,000 home. When you square the errors (as you do when calculating MSE), the difference becomes enormous. 

But since both errors are 5\%, maybe you want to penalize them the same. One option is to use Mean Percentage Error (MPE), but this has the weird effect that if you over-predict one home by 5\% and under-predict the other by 5\%, your MPE is zero. Therefore, a popular option is to use the Mean Absolute Percentage Error (MAPE), which is the mean of the absolute values of the percentage errors and thus would be 5\% in this example. 

For many prediction problems in the policy sphere, we may not only care about accuracy of prediction but also about fairness or other  objectives. The loss function is a place where we can explicitly tell the model to optimize for these concerns in addition to predictive performance.

## K-Nearest Neighbors

### Concept
The idea underlying K-Nearest Neighbors (KNN) is that we expect observations with similar features to have similar outcomes. KNN makes no other assumptions about functional form, so it is quite flexible. 

### Method
KNN can be used for either regression or classification, though it works slightly differently depending on what setting we are in. In the classification setting, the prediction is a majority vote of the observation's $K$-nearest neighbors. In the regression setting, the prediction is the average outcome of the observation's $K$-nearest neighbors. 

For KNN, bias will be lower when $K$ is lower. Bias will increase quickly as k increases, with further away neighbors being included in the prediction.

The only choice we have to make when implementing KNN is the value of $K$ (e.g. how many neighbors should we use in our prediction?). A good way to find $K$ is through cross-validation, something we will cover a little later, but which broadly involves training the algorithm on one set of data and seeing how well it does on a different set.

### Implementation and Considerations
A concern with KNN is whether you have good coverage of your feature space. Imagine that all of your training points were in one region of the feature space, but some of your test points are far away from this region. You will still use the $K$ nearest neighbors to predict the outcome for these far-away test points, but it might not work as well as if the points were close together. Therefore, when implementing KNN, it's good to think about how similar the features in your test set will be to the features in your training set. If they differ systematically, that is a concern (as it would be for other ML methods as well).

Another important consideration is whether there is an imbalance in the frequency of one outcome compared to another. For example, suppose we are trying to classify points as ``true'' or ``false'' and most points are ``true.'' Even if the ``false'' outcomes are clustered together in the feature space, if we use a large enough value of $K$, we will predict ``true'' for these observations simply because there are many more ``true'' observations than ``false'' observations. Therefore, we would do better to use a small value for $K$ in this setting.

Another consideration is whether proximity in each variable is equally important or if proximity in one variable is more important than proximity in another variable. KNN will normalize variables so that they are all on the same scale (same mean and variance) and then treat distance in all normalized variables the same. If you want to up-weight proximity for some variables and down-weight it for others, you can change the way each variable is normalized to accomplish this. Alternatively, you can include only those variables you think are important. When you have this type of uncertainty, there are more principled ways of selecting variables that will be discussed later in the course.

### Extensions 
You might think that neighbors that are really close should be weighted more than neighbors that are a bit further away. Many people agree, so there are methods to allow you to weight different observations differently. You might also think that you shouldn't use just the $K$ nearest neighbors, but all the neighbors within a certain distance. Or maybe you think there's information available in all observations, but there's more information in closer neighbors. All of these adjustments fall under the umbrella of **kernel regression**. In fact, KNN is a special case of kernel regression. Broadly defined, kernel regression methods are a class of methods that generate predictions by taking weighted averages of observations. Because these methods (KNN included) do not specify a functional form, they are called ``non-parametric regression'' methods.


## Linear Regression

### Concept
Linear regression is a parametric model that is additive and linear in the provided features. It is a classic technique used in many fields, and its widespread popularity greatly pre-dates the popularity of machine learning. Its general form is

\begin{equation}
    \hat{y} = \hat{\beta}X
\end{equation}

Where $\hat{y}$ is a vector of predicted $y$ values and $X$ is a matrix whose rows correspond to observations and whose columns correspond to features.

When there is only one feature on the right hand side, the model is called a ``simple linear regression." When there are multiple features on the right hand side, the model is called ``multiple linear regression."

When used for inference, we are interested in $\hat{\beta}$. However, when used for prediction, we are only interested in $\hat{y}$, and we cannot say that the $\hat{\beta}$s reflect any sort of causal relationship between the features and the outcome. For more information on how to test the significance of regression coefficients, please see Chapter 3 of *ISLR* for a reference on $t$-tests (in the simple model) and $F$-tests (in the multivariate model). 

### Method 
To find the coefficients $\hat{\beta}$ in a linear regression, we find the value of $\hat{\beta}$ that minimizes the residual sum of squares (RSS) in the training data. The classic formula for $\hat{\beta}$ uses matrix algebra and is
\begin{equation}
    \hat{\beta} = (X^\prime X)^{-1}X^\prime y
\end{equation}
We will estimate $\hat{\beta}$ using statistical software. 

It is worth noting that the traditional measure of fit for linear regression is $R^2$, but $R^2$ mechanically increases with the inclusion of additional features. Therefore, in the prediction setting, the $R^2$ on the training data is less important than the mean squared error (MSE) on the test data.


### Implementation and Considerations

There are a few things to watch out for as far as the features that you feed into a linear regression. 


- There must be fewer features than observations. Later in the semester, we will cover penalized regression methods that do variable selection to yield estimable linear models, even when the number of available features exceeds the number of observations. Common penalized regression methods are lasso and ridge regression.

- You can use quantitative or qualitative features for the $X$s. When using qualitative features, generate indicator variables for all but one category. The omitted category will serve as the ``baseline,'' meaning that the coefficients on the included categories can be thought of as the differential effect of being in that category compared to the baseline (omitted) one.

- The reason you omit one category when making indicator variables is to avoid linear dependence. If all categories were represented, the indicator columns would all sum to 1, which would mean they were linearly dependent. More generally, you cannot have collinearity or multi-collinearity, which means you cannot have features that are (close to) perfectly correlated.

- You can interact two features (e.g. create a feature that is the product of two other features), and such interactions are valid on categorical and continuous features. However, when you include an interaction, you should also include each of the features on their own as well. Interactions have intuitive appeal if you think there are synergies between two features in terms of their effect on $y$.

- You can exponentiate features and include the exponentiated features in your model. The resulting model is sometimes called polynomial regression and is appropriate when there appears to be a non-linear relationship between a feature and the outcome. 

- Check for influential points -- those that are both *outliers* (they have an unusual or extreme $y$ value) and *high leverage* (they have an unusual or extreme $x$), as these points can greatly influence the model fit. You may want to exclude them or at least check your model's sensitivity to including them versus excluding them.


If you are interested in inference (e.g. looking at the $\hat{\beta}$s to understand a causal relationship), it is important to be aware of Omitted Variables Bias (OVB). OVB occurs when you have two correlated features that each have an effect on $y$ but only one is included in the regression. In that case, the coefficient on the included feature is biased, because it is partially picking up the true effect of the feature on the outcome and is also partially picking up the effect of the omitted feature on the outcome (since the omitted feature is correlated with the included feature). 

As an example of OVB, suppose $X_1$ and $X_2$ are positively (but not perfectly) correlated. If they are also both positively correlated with $y$, then when $X_2$ is omitted from the regression, the coefficient on $X_1$ will be higher than when both $X_1$ and $X_2$ are included. This is because the coefficient on $X_1$ will now pick up both the effect of $X_1$ on $y$ and part of the effect of $X_2$ on $y$ (since $X_1$ is a proxy for $X_2$ because the two features are positively correlated).

When implementing linear regression, you should also take a look at your residuals and make sure there are no red flags:

- When you plot residuals, they should appear randomly scattered. Any non-linearity or patterns in the residuals suggest your model is not appropriate.

- Linear regression assumes residuals are uncorrelated. Evidence of correlated residuals indicates a problem with your model or your data that should be investigated.

- Residuals should have constant variance. If you plot your residuals and their variance seems to be a function of $x$, then the errors are *heteroskedastic* (a fancy word for ``a function of $x$''). In this case, traditional statistical measures of significance are invalid, but other valid methods are available.


 

### Comparison to KNN
The main difference of note between linear regression and KNN is that linear regression is a parametric model whereas KNN is a non-parametric model. There are a few general differences between parametric and non-parametric models that are worth noting

- Non-parametric models are more flexible whereas parametric models impose stronger assumptions

- When there is a small number of observations per feature, parametric models tend to outperform non-parametric models 

In addition to the general differences between parametric and non-parametric models, a key difference between KNN and linear regression is that linear regression is quite simple to fit. In fact, it only requires estimation of a few $\beta$s, whereas KNN is much more computationally intensive. Because of this simplicity, linear regression is also more interpretable than KNN.

# Coding Section


There are lots of great datasets available as part of R packages. Page 14 of Introduction to Statistical Learning with Applications in R Table 1.1 lays out 15 data sets available from R packages. We will use the College dataset from the ISLR package. The first time you ever use a package, you need to install it. Then, every time you want to use the package, you use `library(package_name)`. We will use the college data. Note that details on this data are available online: https://cran.r-project.org/web/packages/ISLR/ISLR.pdf Page 5. You can also get the same information in R by typing: help("College") or ?College. 

## Explore Data


```{r}
library(ISLR)

data(College)
college_data  <- College
```

Let's learn about our data. To get the names of the columns in the dataframe, we can use the function `colnames()`

```{r}
colnames(college_data)
```

To find out how many rows and columns are in the dataset, use `dim()` Recall that this gives us Rows followed by Columns

```{r}
dim(college_data)
```

You can also look in the "environment" tab, press the blue arrow next to college_data and it will drop down showing the column names with their types and first few values. For college, all columns except the first are numeric. The first column is a factor column, which means it's categorical. To get a better sense of the data, let's look at it:


```{r, eval=F}
View(college_data)
```

Suppose we are interested in predicting whether a college is private or public based on available covariates, like Number accepted, enrolled, etc. Additionally, let's suppose you don't want certain variables included in your dataset. You can drop these functions using -c(). For example, let's suppose you don't want the Apps or Student to Faculty Ratio included in your dataset.  

```{r}
college_data <- college_data[, -c(15, 2)]
```

Be careful when you are dropping multiple columns. You need to put the numbers in reverse order (from highest to lowest). This is because if you drop the second column first, then the 15th column becomes the the 14th column.

```{r}
college_data <- College
college_data <- college_data[, -c(2)]
college_data <- college_data[, -c(15)]
```

A less manual way of dropping columns is to use R to first use R to find the corresponding indices in the data columns. Go back to the original college data

```{r}
college_data <- College
```

Find the indices (i.e. column positions) of the columns to drop 

```{r}
to_drop <- which(names(college_data) %in% c("Apps", "S.F.Ratio"))
print(to_drop)
```

Reverse the indices as suggested above

```{r}
to_drop <- rev(to_drop)
print(to_drop)
```

Now use the object you have defined to drop the columns

```{r}
college_data <- college_data[, -c(to_drop)]
```


Also sometimes we have factor variables that we want to convert to numeric variables. To check variable types, you can use the "str" function

```{r}
str(college_data)
```

You can see that the Private variable is a factor. We can convert it to a numeric variable using the "as.numeric" function. I like my binary variables in R to be 0/1. In R, most factors automatically convert to a binary 1/2 format. I usually prefer a binary 0/1 format. To transform, I subtract 1. 

```{r}
college_data$Private <- as.numeric(college_data$Private) - 1 
summary(college_data$Private)
summary(College$Private)
```


Let's get back our original sample 
```{r}
college_data <- College
```


## Testing and Training Sets

In order to make this interesting, let's split our data into a training set and a test set. To do this, we will use `set.seed()`, which will allow us to draw the same pseudorandom numbers the next time we run this code, and we will use the `sample()` function. 



```{r}
set.seed(222)
```

The `sample()` function takes two arguments: The first is a vector of numbers from which to draw a random sample. The second is the number of random numbers to draw. The default is to sample without replacement, but you can sample with replacement by adding "`, replace = TRUE`" inside the function. Now, let's generate a list of indices from the original dataset that will be designated part of the test set using `sample()`

```{r}
test_ids <- sample(1:(nrow(college_data)), round(0.2 * nrow(college_data)))
```

To identify the training_ids, we want all of the numbers from 1:nrow(college_data) that aren't test IDs. Recall that `which()` returns the indices for which the statement inside the parentheses is true. `which(!())` returns the indices for which the statement inside the parentheses is false. The "!" means "not". Also, if you wanted to know which values of vector A were in vector B, you can use `which(A %in% B)`. So if you want to know which values of vector A are NOT in vector B, you use `which(!(A %in B))`, so that's what we will do -- vector A is the vector of all integers between 1 and the number of rows in our data. vector B is the vector of test IDs

```{r}
training_ids <- which(!(1:(nrow(college_data)) %in% test_ids))
```


We can use these indices to define our test and training sets by putting those vectors in the row position inside square brackets.

```{r}
test_data <- college_data[test_ids,]
training_data <- college_data[training_ids,]
```

## KNN Classification

Let's develop a KNN model to try to predict whether it's a private college using all available features.

To use KNN for classification, we need to install and load the library "class"
```{r}
library(class)
```

knn() is the function we will use to run the KNN model. It takes four arguments:

- train = training data features (no outcome)
- test = test data features (no outcome)
- cl = training data outcome (class each observation belongs to)
- k = number of nearest neighbors to use

For two-class classification problems, k should be odd (avoids tied votes). Let's run the model with 1 NN and 9 NNs. To exclude a column, use -# in the column position insider square brackets. (e.g. df[, -2] excludes the second column of dataframe df)

```{r}
knn_model1 <- knn(train = training_data[, -1],
                  test = test_data[, -1],
                  cl = training_data[, 1],
                  k = 1)

knn_model9 <- knn(train = training_data[, -1],
                  test = test_data[, -1],
                  cl = training_data[, 1],
                  k = 9)
```


We are trying to predict Private Yes/No. `knn()` output predicted values for our test data, so we can compare actual v. predicted values. "prediction == actual" gives a vector with the same number of elements as there are observations in the test set. Each element will either be TRUE (the prediction was correct) or FALSE (the prediction was wrong). Applying which() to this vector will yield the index numbers for all the elements equal to TRUE. Applying length() to that vector tells us how many are TRUE (e.g. for how many observations prediction == actual). We can then divide by the number of observations in the test data to obtain the accuracy rate

```{r}
accuracy1  <- length(which(knn_model1 == test_data$Private)) / nrow(test_data)
accuracy9 <- length(which(knn_model9 == test_data$Private)) / nrow(test_data)

print(accuracy1)
print(accuracy9)
```

Let's visualize what is happening in a KNN classification model. We will use the `ggplot2` package to create a scatterplot of the training data, and then overlay the test data on top of it. We will color the points by whether the school is private or not.

```{r}
library(ggplot2)
ggplot(data = training_data, 
       aes(x = Outstate, y = F.Undergrad, 
           color = as.factor(Private))) +
  geom_point() +
  geom_point(data = test_data, aes(x = Outstate, y = F.Undergrad), 
             color = "black", size = 1) +
  scale_color_manual(values = c("red", "blue")) +
  theme_minimal() +
  guides(color = guide_legend(title = "Private")) 

```


This seems like excellent predictive performance. However, it's good to think about the distribution of the data. As an extreme example, if all schools in the data were private, we would expect 100% prediction accuracy regardless of our model. Let's see how well we do if our prediction is all schools are Private. Start by calculating the proportion of private schools

```{r}
print(length(which(test_data$Private == "Yes")) / nrow(test_data))
```

We can also check our accuracy on Private schools v. Public schools. To do this, we need to figure out which schools are private in the test data. Specifically, get the indices for the private schools

```{r}
private_schools <- which(test_data$Private == "Yes")
public_schools <- which(test_data$Private == "No")

print(private_schools)
print(public_schools)
```

To calculate the prediction accuracy for private schools, we need to know how many (true not predicted) private schools are in the test data. Likewise, we need to know how many public schools are in the test data.

```{r}
num_private_schools <- length(private_schools)
num_public_schools <- length(public_schools)
```

Now we will calculate the prediction accuracy separately for private and public schools. 

```{r}
private_accuracy1 <- length(
  which(knn_model1[private_schools] == test_data$Private[private_schools])) /
  num_private_schools

private_accuracy9 <- length(
  which(knn_model9[private_schools] == test_data$Private[private_schools])) /
  num_private_schools
```

Now we will calculate the prediction accuracy separately for private and public schools. 

```{r}
## Private schools (% correctly predicted):
private_accuracy1 <- length(
  which(knn_model1[private_schools] == test_data$Private[private_schools])) /
  num_private_schools

private_accuracy9 <- length(
  which(knn_model9[private_schools] == test_data$Private[private_schools])) /
  num_private_schools
```

```{r}
# Public schools (% correctly predicted): 
public_accuracy1 <- length(
  which(knn_model1[public_schools] == test_data$Private[public_schools])) /
  num_public_schools

public_accuracy9 <- length(
  which(knn_model9[public_schools] == test_data$Private[public_schools])) /
  num_public_schools
```


Let's see how it did on different school types:
```{r}
print(private_accuracy1)
print(public_accuracy1)

print(private_accuracy9)
print(public_accuracy9)
```


Therefore, we did better on private schools than public schools because our prediction accuracy was higher on private schools. Thinking about differential performance by label is related to fairness of machine learning algorithms. For an interesting discussion on ML fairness and different ways to define fairness, see the following academic paper: 

    Jon Kleinberg, Sendhil Mullainathan, and Manish Raghavan. 
    Inherent Trade-Offs in the Fair
    Determination of Risk Scores, November 2016


## KNN for Regression

Suppose we wanted to predict how many students would enroll given the other features available in the data. In that case, the classification function we used above will not work. We will need a KNN function designed for regression problems. This function is `knn.reg()` in the `FNN` package, so we should install then read in the FNN package.

```{r, message=FALSE, warning=FALSE}
#install.packages("FNN")
library(FNN)
```


knn.reg() takes four arguments:
- training data with only features (no outcome)
- test data with only features (no outcome)
- training outcomes
- k = number of neighbors

Enrollment is the fourth column, so we will exclude that from the features. Because public / private is a factor, we either need to convert it to a numeric variable or exclude it. We will exclude it for now. Note that you can scale your features using `scale()`. Deciding to scale your features or not is problem dependent. We will not scale here. If you're not sure whether or not to scale, you can always try it both ways and see how the performance changes.


```{r}
knn_reg1 <- knn.reg(training_data[, -c(1, 4)],
                    test_data[, -c(1, 4)],
                    training_data$Enroll,
                    k = 1)

knn_reg5 <- knn.reg(training_data[, -c(1, 4)],
                    test_data[,-c(1, 4)],
                    training_data$Enroll,
                    k = 5)
```



MSE is an appropriate loss function for regression whereas accuracy is only relevant for classification

```{r}
mse_knn1 <- mean((knn_reg1$pred - test_data$Enroll)^2)
mse_knn5 <- mean((knn_reg5$pred - test_data$Enroll)^2)

print(mse_knn1)
print(mse_knn5)
```

## Standard Linear Regression

We will now do linear regression. To run a linear regression in R, we use the function `lm()`, which stands for linear model. `lm()` takes two main arguments. The first is the formula, which should be of the form Dependent Variable ~ Feature1 + Feature2 + ... The second is the training data -- including both features and the outcome. Note that "`~.`" means regress this variable on all other variables

```{r}
enroll_reg <- lm(Enroll ~ ., training_data)

```

lm() returns a list, which includes among other things coefficients, residuals, and fitted values for the training data. You can look at the elements in RStudio by using the blue arrow next to enroll_reg in the environment tab. In order to call one element of a list, you can use $

```{r}
enroll_reg$coefficients
```


In order to see a more traditional regression output, use `summary()`

```{r}
summary(enroll_reg)
```


If you want to use the coefficients from enroll_reg to predict enrollment values in the test data, you can use the function `predict()`. The first argument is the lm object (the whole thing -- not just the coefficients) and the second argument is the test data frame without the outcome column

```{r}
predicted_enroll <- predict(enroll_reg, test_data[, -4])
```

Let's see how well we did in terms of MSE

```{r}
MSE_lm_enroll <- mean((predicted_enroll - test_data$Enroll)^2)
print(MSE_lm_enroll)
```

We can see how this compared to our training MSE

```{r}
print(mean((enroll_reg$residuals)^2))
```

Training MSE as % of Test MSE:

```{r}
print(mean((enroll_reg$residuals)^2) / MSE_lm_enroll)
```


We know that the coefficients might change if we exclude some variables. Let's pretend we only had Apps and Accept (columns 2 and 3) as features

```{r}
small_enroll_reg  <- lm(Enroll ~ Apps + Accept, training_data)
```

We can compare coefficients from the small regression and the full regression. If the coefficients in the small regression are different from the coefficients in the full regression, then the small regression suffers from Omitted Variables Bias (OVB).

```{r}
small_enroll_reg$coefficients
enroll_reg$coefficients
```


## Stargazer for Regression Output

If you want to compare the coefficients from different regressions, you can use the `stargazer` package. This package is not installed by default, so you will need to install it. 

```{r, message=FALSE, warning=FALSE}
#install.packages("stargazer")
library(stargazer)

stargazer(small_enroll_reg, enroll_reg, 
          type = "text", column.labels = c("Small Model", "Full Model"))
```

You can also also use stargazer to get the latex code for a table. This is useful if you want to include the table in a paper or a presentation. 

```{r, message=FALSE, warning=FALSE, eval=FALSE}
stargazer(small_enroll_reg, enroll_reg, 
          type = "latex", 
          column.labels = c("Small Model", "Full Model"))
```

<table style="text-align:center"><tr><td colspan="3" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left"></td><td colspan="2"><em>Dependent variable:</em></td></tr>
<tr><td></td><td colspan="2" style="border-bottom: 1px solid black"></td></tr>
<tr><td style="text-align:left"></td><td colspan="2">Enroll</td></tr>
<tr><td style="text-align:left"></td><td>Small Model</td><td>Full Model</td></tr>
<tr><td style="text-align:left"></td><td>(1)</td><td>(2)</td></tr>
<tr><td colspan="3" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">PrivateYes</td><td></td><td>7.807</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(30.036)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Apps</td><td>-0.052<sup>***</sup></td><td>-0.028<sup>***</sup></td></tr>
<tr><td style="text-align:left"></td><td>(0.013)</td><td>(0.008)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Accept</td><td>0.424<sup>***</sup></td><td>0.147<sup>***</sup></td></tr>
<tr><td style="text-align:left"></td><td>(0.021)</td><td>(0.015)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Top10perc</td><td></td><td>4.016<sup>***</sup></td></tr>
<tr><td style="text-align:left"></td><td></td><td>(1.283)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Top25perc</td><td></td><td>-2.269<sup>**</sup></td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.997)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">F.Undergrad</td><td></td><td>0.144<sup>***</sup></td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.004)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">P.Undergrad</td><td></td><td>-0.012<sup>*</sup></td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.007)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Outstate</td><td></td><td>-0.003</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.004)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Room.Board</td><td></td><td>-0.024<sup>**</sup></td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.011)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Books</td><td></td><td>-0.027</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.049)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Personal</td><td></td><td>0.008</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.014)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">PhD</td><td></td><td>-0.431</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(1.002)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Terminal</td><td></td><td>-0.540</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(1.094)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">S.F.Ratio</td><td></td><td>-0.253</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(2.843)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">perc.alumni</td><td></td><td>2.319<sup>***</sup></td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.879)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Expend</td><td></td><td>0.003</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.003)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Grad.Rate</td><td></td><td>0.136</td></tr>
<tr><td style="text-align:left"></td><td></td><td>(0.648)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td style="text-align:left">Constant</td><td>86.881<sup>***</sup></td><td>187.938<sup>**</sup></td></tr>
<tr><td style="text-align:left"></td><td>(20.984)</td><td>(89.615)</td></tr>
<tr><td style="text-align:left"></td><td></td><td></td></tr>
<tr><td colspan="3" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">Observations</td><td>622</td><td>622</td></tr>
<tr><td style="text-align:left">R<sup>2</sup></td><td>0.820</td><td>0.956</td></tr>
<tr><td style="text-align:left">Adjusted R<sup>2</sup></td><td>0.819</td><td>0.955</td></tr>
<tr><td style="text-align:left">Residual Std. Error</td><td>403.989 (df = 619)</td><td>202.349 (df = 604)</td></tr>
<tr><td style="text-align:left">F Statistic</td><td>1,405.761<sup>***</sup> (df = 2; 619)</td><td>768.822<sup>***</sup> (df = 17; 604)</td></tr>
<tr><td colspan="3" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left"><em>Note:</em></td><td colspan="2" style="text-align:right"><sup>*</sup>p<0.1; <sup>**</sup>p<0.05; <sup>***</sup>p<0.01</td></tr>
</table>
