---
title: "Reproducibility in Health Research - The Power of Docker for Microsimulations"
description: "🐳🐳🐳"
author: "Jacob Jameson"
date: "2023-09-23"
categories: [reproducibility, docker]
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
link-citations: yes
reference-location: margin
comments:
  utterances:
    repo: jacobjameson/jacobjameson.github.io
format:
  html:
    theme: slate
---

```{r setup, include=FALSE}
library(tufte)
library(tidyverse)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

## Introduction
In the domain of health research, particularly within the realm of medical decision-making, our ability to make accurate and impactful decisions is profoundly influenced by the robustness and reliability of our research. However, robustness is not just about good statistical practices or rigorous study design; it extends to the reproducibility of our research. Without reproducibility, even the most meticulously designed studies can lose their value and, in some cases, credibility.

### The Imperative of Reproducible Science
Medical decision-making often hinges on intricate computational models such as microsimulations. These models, while powerful, are composed of multifaceted components including various dependencies, data sources, and algorithms. A slight change or inconsistency in any of these components can lead to vastly different outcomes.

Imagine, for instance, a microsimulation that forecasts the spread of an infectious disease. Policy-makers could utilize this simulation to shape national health strategies. Now imagine if a slight variation in the software environment—say a minor version change in a data processing library—alters the projected number of cases. The implications could be enormous, ranging from resource misallocation to ineffective interventions.

This underlines a fundamental truth: For research to be trusted and actionable, it must be reproducible.

### The Container Solution: Demystifying Docker
So, how can we ensure that our complex microsimulations run consistently, producing the same outcomes regardless of where and when they are executed? Enter Docker, a game-changer in the world of reproducible research.

#### What is a Container?

Think of a container as a standalone package, a box if you will, that encloses your entire software environment—your code, its dependencies, system libraries, system settings, and so forth. This container ensures that your software runs identically regardless of where the container is executed, be it your laptop, a colleague's machine, or a cloud server on the other side of the world.

Why is this groundbreaking? Because it eradicates the notorious "but it works on my machine" problem. Every researcher knows the headache of trying to replicate an environment across multiple machines, battling inconsistent library versions, missing dependencies, and unexpected system behaviors. Docker containers eliminate this inconsistency.

### Docker and Reproducibility

With Docker, every element of your software environment is codified and encapsulated. You don't just share your code; you share its entire ecosystem. Here's what it means for reproducibility:

1. Isolation: Docker ensures that your software's environment is isolated from the host system, meaning external factors on a machine won't affect your software's execution.


2. Version Control for Environments: Just as Git allows you to version control your code, Docker lets you version control your software environment. This is crucial when using libraries and tools that are continuously updated.

3. Ease of Sharing: Docker containers can be easily shared with peers, ensuring that they don't just get your code but the exact environment in which your code was designed to run.

As health researchers, our ultimate aim is to produce knowledge that can improve health outcomes. But this knowledge must be built on trust. Docker, by championing reproducibility, fosters this trust, ensuring our research remains both impactful and credible. In the next sections, we'll delve deeper into how to leverage Docker for your microsimulations, but for now, it's crucial to recognize its potential in safeguarding the integrity of our research.


## Building a Reproducible Microsimulation with Docker

1. Setting Up the Microsimulation Project

Firstly, we'll start by creating the microsimulation itself. Let's use the stochastic infectious disease model we discussed earlier.

Create a new directory for your project:

```{r, eval=F}
mkdir microsim_disease_model
cd microsim_disease_model
```

In this directory, you'll have:

1. A script named microsimulation.R containing the microsimulation code.
2. Any additional data or configuration files your simulation might requir

