---
title: "Section 7 - Non-linear Models"
from: markdown+emoji
format:
  html:
    toc: true
    toc-location: left
page-layout: full
description: |
author: Jacob Jameson

---


Note: this section has several different types of models. We cannot cover all of them in 1.25 hours. We will go over a few examples but once you understand how the examples, you should be able to apply the logic to other types models. 


This week, we will use the Wage data that is part of the ISLR package

```{r}
library(ISLR)
```


Load the Wage data and drop the "Wage" columns, as we did in Section 6

```{r}
wage_data <- Wage
wage_data <- wage_data[, -10]
```


  
To make life easier for later analyses, I will start by sorting the data on age. To do this, I will use a package called `dplyr`, which has a lot of great tools for data manipulation. 
  
```{r, message=FALSE, warning=FALSE}
library(dplyr)
```
  
The first two functions we will use are:
- `%>%` which means "and then do"
- `arrange()` which sorts the data on the column inside the parentheses

```{r}
wage_data <- wage_data %>% arrange(age)
```


We will start by splitting the data into training and test sets

```{r}
set.seed(222)
train <- sample(1:nrow(wage_data), round(nrow(wage_data) * 0.8))
train <- sort(train)
test <- which(!(seq(nrow(wage_data)) %in% train))
```

To quickly and easily measure MSEP, write our own function

```{r}
msep_func <- function(predictions, true_vals) {
  MSEP <- mean((predictions - true_vals)^2)
  return(MSEP)
}
```


## Polynomial Regression 

We can start by fitting a polynomial regression using only age

```{r}
age_poly <- lm(wage ~ poly(age, 4), data = wage_data[train,])
```


Extract the coefficients from the model

```{r}
coef(summary(age_poly))
```

When you use `poly()`, it returns a matrix of "orthogonal polynomials" so the columns of the matrix are linear combinations of age, age$^2$, age$^3$, age$^4$. Let's take a look!

```{r}
head(poly(wage_data$age, 4))
head(wage_data$age)
```

If you want it to return the raw powers of age, you can add an argument to the function `poly()`

```{r}
head(poly(wage_data$age, 4, raw = TRUE))
```

Although the two forms give you different numbers, they result in the same predictions, because your model is still a linear combination of the original powers

```{r}
age_poly_TRUE <- lm(wage ~ poly(age, 4, raw = TRUE), 
                    data = wage_data[train,])
```

Extract the coefficients from the model

```{r}
coef(summary(age_poly_TRUE))
```

We can see from the coefficient outputs of the two models that they have different coefficients. We will now check that they make the same predictions.

```{r}
age_poly_pred <- predict(age_poly, newdata = wage_data[test,])
age_poly_TRUE_pred  <- predict(age_poly_TRUE, newdata = wage_data[test,])
```

Explore the predictions

```{r}
head(age_poly_pred)
head(age_poly_TRUE_pred)
```

Calculate and print the MSEP

```{r}
print(msep_func(age_poly_pred, wage_data[test, "wage"]))
```


`cbind()` is a function that joins columns together by binding them next to each other. There is also a function `rbind()` that joins by binding new rows to the bottom of old ones. Let's try using `cbind()` to look at both sets of predictions at the same time:

```{r}
pred_comparison <- cbind(age_poly_pred, age_poly_TRUE_pred)
```


The column names default to the variable names, but ifwe want to change them, we can use the `colnames()` function

```{r}
colnames(pred_comparison) <- c("pred1", "pred2")
```

We can then generate a variable that flags any instances where the predictions do not line up. To do this, I will introduce you another function in `dplyr`: `mutate()` which means create a new variable

```{r, eval=FALSE}
pred_comparison <- pred_comparison %>% 
  mutate(check_same = as.numeric(pred1 == pred2))
```
`Error in UseMethod("mutate") : no applicable method for 'mutate' applied to an object of class "c('matrix', 'array', 'double', 'numeric')"`

The previous line errored because `pred_comparison` is a matrix and dplyr was designed to work on data frames.

Make `pred_comparison` a data frame

```{r}
pred_comparison <- data.frame(pred_comparison)
```

Now try running the code:

```{r}
pred_comparison <- pred_comparison %>%
  mutate(check_same = as.numeric(pred1 == pred2))
```

View the results

```{r, eval=F}
View(pred_comparison)
```

## Regression Splines

### Splines

We're going to start by making an age grid, so that we test our predictions on a grid of evenly spaced ages so that you capture the functional form well. 

A quick aside, na.rm = TRUE means exclude missing observations. Then we are going to run a spline with knots at ages 25, 40, and 60. 

```{r}
library(splines)
```


Create the age grids

```{r}
age_grid <- seq(from = min(wage_data$age, na.rm = TRUE),
                to = max(wage_data$age, na.rm = TRUE))
```

Now we use a basis function

```{r}
spline_age <- lm(wage ~ bs(age, knots = c(25, 40, 60)), 
                 data = wage_data[train,])

?bs # to check what the basis function does
```


Get the predictions at the grid points we defined earlier

```{r}
spline_age_grid_pred  <- predict(spline_age, 
                                 newdata = list(age = age_grid), 
                                 se = TRUE)
```

Plot age on the x-axis and wage on the y-axis for the test data. Then add the predictions and the confidence intervals

```{r}
plot(wage_data[test, "age"], wage_data[test, "wage"],
cex = 0.5, col = "darkgrey",
xlab = "age", ylab = "wage")
lines(age_grid, spline_age_grid_pred$fit, lwd = 2)
lines(age_grid, spline_age_grid_pred$fit + 
        2 * spline_age_grid_pred$se, lty ="dashed")
lines(age_grid, spline_age_grid_pred$fit - 
        2 * spline_age_grid_pred$se, lty ="dashed")
# add vlines for knots
abline(v = c(25, 40, 60), lty = "dotted")
```

### Natural Splines 

If we instead wanted to fit a natural spline, we use `ns()` instead of knots, we can specify degrees of freedom. In this case we are going to use 4 degrees of freedom. 

```{r}
ns_age_poly <- lm(wage ~ ns(age, df = 4), data = wage_data[train,])
ns_age_grid_poly_pred <- predict(ns_age_poly,
                                 newdata = list(age = age_grid),
                                 se = TRUE)

plot(wage_data[test, "age"], wage_data[test, "wage"], 
     cex = 0.5, col = "darkgrey",
     xlab = "age", ylab = "wage")
lines(age_grid, ns_age_grid_poly_pred$fit, lwd = 2)
lines(age_grid, ns_age_grid_poly_pred$fit + 
        2 * ns_age_grid_poly_pred$se, lty = "dashed")
lines(age_grid, ns_age_grid_poly_pred$fit - 
        2 * ns_age_grid_poly_pred$se, lty = "dashed")
abline(v = quantile(wage_data[train,]$age, 
                    probs = c(0.25, 0.5, 0.75)), lty = "dotted")
```

### Smoothing Splines 

To fit a smoothing spline, we use `smooth.spline()`. We can specify our own `df`   

```{r}
smooth_age <- smooth.spline(wage_data[train,"age"], 
                            wage_data[train, "wage"], 
                            df = 16)
```

Or we can use cross Validation to get optimal df and penalty   

```{r, warning=FALSE}
smoothCV_age <- smooth.spline(wage_data[train, "age"], 
                              wage_data[train, "wage"], 
                              cv = TRUE) # specify we want to use CV
smoothCV_age$df
```

Plot the results as before  

```{r}
plot(wage_data[test, "age"], wage_data[test, "wage"],
     cex =.5, col = "darkgrey",
     xlab = "age", ylab = "wage")
title(" Smoothing Spline ")
lines(smooth_age, col ="red", lwd = 2)
lines(smoothCV_age, col ="blue", lwd =2)
```

### Local Regression 

Local Regression use loess()

Note: span = 0.2 makes neighborhoods with 20% of observations. Span = 0.5 creates neighborhoods with 50% of observations. So the larger the span, the smoother the fit

```{r, warning=F, message=F}
# plot age and wage and add geom_smooth
library(ggplot2)
ggplot(wage_data[test,], aes(x = age, y = wage, color = education)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = T, color = "black") +
  #geom_smooth(method = "loess", span = 0.2) +
  geom_smooth(method = "loess", span = 0.5) +
  facet_wrap(~education) 
```


```{r}
local2_age <- loess(wage ~ age, span = 0.2, data = wage_data)
local5_age <- loess(wage ~ age, span = 0.5, data = wage_data)
```

Get the predictions

```{r}
pred_local2_age <- predict(local2_age, newdata = data.frame(age = age_grid))
pred_local5_age <- predict(local5_age, newdata = data.frame(age = age_grid))
```

Plot the results

```{r}
plot(wage_data[test, "age"], wage_data[test, "wage"],
     cex =.5, col = "darkgrey",
     xlab = "age", ylab = "wage")
lines(age_grid, pred_local2_age, col = "red", lwd = 2)
lines(age_grid, pred_local5_age, col = "blue", lwd = 2)
```

### GAMs 

Generalized Additive Models (GAMs)

Start with using natural spline functions of year and age and treating education as a qualitative variable does not require any special packages

- Example 1: GAMs with splines

```{r}
gam_yae <- lm(wage ~ ns(year, 4) + ns(age, 4) + education, 
              data = wage_data[train,])
```

- Example 2: GAMs with smoothing splines

```{r, message=FALSE}
library(gam)
```

In the gam library s() indicates that we want to use a smoothing spline

```{r}
gam_smooth <- gam(wage ~ s(year, 4, spar = 0.5) + s(age, 5) + education,
                  data = wage_data[train,])
```

Plot the model results

```{r}
par(mfrow = c(1, 3))
plot(gam_smooth, se = TRUE, col ="blue ")
```

To plot the GAM we created just using `lm()`, we can use `plot.Gam()`
```{r}
plot.Gam(gam_yae, se = TRUE, col = "red") # Note the capitalization
```

Make predictions

```{r}
gam_yae_pred <- predict(gam_yae, newdata = wage_data[test,])
gam_smooth_pred <- predict(gam_smooth, newdata = wage_data[test,])
```

Print out the MSEP for these two GAMs using the function we created at the start of class

```{r}
print(msep_func(gam_yae_pred, wage_data[test, "wage"]))
print(msep_func(gam_smooth_pred, wage_data[test, "wage"]))
```

- Example 3: GAMs with local regression

To use local regression in GAMs, use `lo()`

```{r}
gam_lo <- gam(wage ~ s(year, df = 4) + lo(age, span = 0.7) + education, 
              data = wage_data[train,])
```

Plot the results

```{r}
plot.Gam(gam_lo, se = TRUE, col = "green")
```

If you want to do a local regression in two variables:

```{r}
gam_2lo <- gam(wage ~ lo(year, age, span =0.2) + education,
               data = wage_data[train,])
```



  
